Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-04-09 manuscript.tex
almost there.

\subsection*{Genotyping Analysis} \jri{please drop scripts for bayenv and STRUCTURE into githu

@@

We genotyped 247 accessions of maize landraces, and found the Hopscotch element fixed in all but 8 of these accessions (\ref{TableS1Mays}, \ref{TableS2}). For all analyses we only included individuals that clearly resolved for both PCRs, resulting in 837 individuals total. Within our \emph{parviglumis} and \emph{mexicana} samples we found the Hopscotch insertion segregating in 37 and 4 populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico in both subspecies (\ref{Fig1Map}). We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p<0.05).

%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig1Map.pdf}
    \caption{Map showing the frequency of the Hopscotch allele in populations of \emph{Zea mays} ssp. \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of alleles sampled. } 
\label{Fig1Map}
  \end{center}
\end{figure*}
%--------------------------------------------

Using our Hopscotch genotyping data, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled 8 or more alleles. F\textsubscript{CT} is 0 within our dataset, and we found similar levels of F\textsubscript{ST} between populations within each subspecies (0.22) and between all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi2013} (\ref{Table1Fst}). 
% Table 1
\begin{table}[htbp]
  \centering
  \caption{Pairwise F\textsubscript{CT} values from sequence data}
    \begin{tabular}{rccc}
 %   \toprule
    \multicolumn{1}{c}{\textbf{Comparison}} & \textbf{Sequenced Region 2} & \textbf{Sequenced Region 1} & \textbf{Hopscotch} \\
 %   \midrule
    EjuA \& EJuB & 0     & 0     & 0 \\
    EjuA \& MSA & 0.328 & 0.326 & 0.186 \\
    EjuA \& SLO & 0.258 & 0.416 & 0.28 \\
    EjuB \& MSA & 0.365 & 0.397 & 0.188 \\
    EjuB \& SLO & 0.29  & 0.512 & 0.28 \\
    MSA \& SLO & 0     & 0.007 & 0.016 \\
%    \bottomrule
    \end{tabular}
  \label{Table1Fst}
\end{table}
Although we found large variation in Hopscotch allele frequency among our populations, our BayEnv analysis did not indicate a correlation between the Hopscotch insertion and environmental variables (all Bayes Factors $<$ 1; \ref{TableS3BayEnv}).

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A has the highest values of haplotype diversity, $\theta_W$, and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is similarly negative in the two Ejutla populations and La Mesa, but is more positive in San Lorenzo (\ref{Table2Diversity}). 
% Table 2 Diversity Statistics
\begin{table}[htbp]
  \centering
  \caption{Add caption}
    \begin{tabular}{rccccc}
%    \toprule
    Population & Number of Haplotypes & Haplotype Diversity & $\theta_W$    & $\theta_\pi$    & Tajima's D \\
%    \midrule
    \multicolumn{6}{c}{\textit{Sequenced region 2 (50kb upstream)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.89394 & 0.01548 & 0.01763 & 0.6231 \\
    \multicolumn{1}{c}{EJUB} & 8     & 0.89394 & 0.01493 & 0.01591 & 0.29504 \\
    \multicolumn{1}{c}{MSA} & 3     & 0.68182 & 0.01111 & 0.01055 & -0.22212 \\
    \multicolumn{1}{c}{SLO} & 4     & 0.74242 & 0.01167 & 0.01413 & 0.93185 \\
    \multicolumn{6}{c}{\textit{Sequenced region 1(5? UTR)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.85897 & 0.00874 & 0.00527 & -1.64955 \\
    \multicolumn{1}{c}{EJUB} & 5     & 0.70909 & 0.00663 & 0.00378 & -1.83123 \\
    \multicolumn{1}{c}{MSA} & 6     & 0.68182 & 0.00646 & 0.00373 & -1.75506 \\
    \multicolumn{1}{c}{SLO} & 3     & 0.31818 & 0.00176 & 0.00137 & -0.72873 \\
    \bottomrule
    \end{tabular}
  \label{Table2Diversity}
\end{table}

For region 2, haplotype diversity, $\theta_{\textsubscript{W}}$, and $\theta_{\textsubscript{\pi}}$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (\ref{Table2Diversity}). Tajima's D is positive in all populations except San Lorenzo, indicating an excess of low frequency variants in this latter population (\ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are 0 for both sequenced regions as well as for the Hopscotch, while they are high for other population pairs (\ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILS; data from Maize HapMapV2, \citep{Chia2012}) do not reveal any clear clustering pattern with respect to population or Hopscotch genotype (\ref{FigureS3NJ}). Furthermore, individuals within our sample that have the Hopscotch insertion do not group with the teosinte inbred lines or the lines of domesticated maize that have the Hopscotch insertion.

@@

The teosinte populations with the highest frequency of the Hopscotch insertion in this study were sympatric with cultivated maize. Our initial hypothesis was that the high frequency of the Hopscotch element in these populations could be attributed to introgression from maize into teosinte. To investigate this possibility we examined overall patterns of linkage disequilibrium across chromosome one, and specifically in the \emph{tb1} region. If the Hopscotch is found in these populations due to recent introgression we would expect to find large blocks of linked markers near this element. We find no evidence of elevated linkage disequilibrium between the Hopscotch and SNPs surrounding the \emph{tb1} region in our resequenced populations (\ref{Fig2LD}), and $r^{2}$ in the \emph{tb1} region (positions 264,596,664-265,891,456; AGPv2) does not differ significantly between populations with (average $r^{2}$ of 0.085) and without the Hopscotch genotype (average $r^{2}=0.082$). In fact, average $r^{2}$ is lower in the \emph{tb1} region ($r^{2}=0.056$) than across the rest of chromosome 1 ($r^{2}=0.083$) (\ref{Table3R2}).

% Table 3 $r^{2}$ values
\begin{table}[htbp]
  \centering
  \caption{$r^{2}$ values between SNPs in the \emph{tb1} region (positions 264,596,664-265,891,456) and the rest of chromosome 1, within the 5? UTR (Sequenced region 1), and within the 66,169 bp upstream region (Sequenced region 2).}
    \begin{tabular}{cccrc}
 %   \toprule
    \multirow{2}[2]{*}{Population} & \multirow{2}[2]{*}{Chromosome 1} & \multirow{2}[2]{*}{tb1 region} & \multicolumn{1}{c}{\multirow{2}[2]{*}{Sequenced region 1}} & \multirow{2}[2]{*}{Sequenced region 1 } \\
 %   \midrule
          &       &       & \multicolumn{1}{c}{} &  \\
    Ejutla A & 0.095426101 & 0.050304 & 0.747295 & 0.214933 \\
    Ejutla B & 0.068681837 & 0.051295 & 0.660354 & 0.186395 \\
    La Mesa & 0.069500533 & 0.053306 & 0.914286 & 0.766234 \\
    San Lorenzo & 0.100536784 & 0.067251 & 0.912281 & 0.636364 \\
    \bottomrule
    \end{tabular}
  \label{Table3R2}
\end{table}
 

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig2LDPlots.pdf}
    \caption{Linkage equilibrium plots for SNPs within bp 261,000,000-268,000,000 on chromosome 1. The yellow rectangle indicates the location of the Hopscotch insertion and the green represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D). San Lorenzo} 
\label{Fig2LD}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

The lack of clustering of Hopscotch genotypes in our NJ tree as well as the lack of LD around \emph{tb1} does not support the hypothesis that the Hopscotch insertion in these populations of \emph{parviglumis} is the result of recent introgression. However, to further explore this hypothesis we performed a STRUCTURE analysis using Illumina MaizeSNP50 data from four of our \emph{parviglumis} populations (EjuA, EjuB, MSA, and SLO) and the maize 282 diversity panel \citep{Cook2012, Phyh\"aj\"arvi2013}. The linkage model implemented in STRUCTURE can be used to identify ancestry of blocks of linked variants, which would arise as a result of recent admixture between populations. If the Hopscotch insertion is present in populations of \emph{parviglumis} as a result of recent admixture with domesticated maize, we would expect the insertion and linked variants in surrounding sites to be assigned to the "maize" cluster in our STRUCTURE runs, not the "teosinte" cluster. In all runs, assignment to maize in the \emph{tb1} region across all four \emph{parviglumis} populations is lower (average 0.017) than the chromosome wide average (0.20; \ref{Fig3Structure}).

% Figure 3 STRUCTURE Plot 4 Pops
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig3Structure.pdf}
    \caption{STRUCTURE assignment to maize across a section of chromosome 1. The dotted lines mark the beginning of the sequenced region 50kb upstream (Sequenced region 2) and the end of the \emph{tb1} ORF. } 
\label{Fig3Structure}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

@@

\subsubsection*{Phenotyping 1}

We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants from 23 sampling locations within the San Lorenzo population, and genotyped plants for the Hopscotch insertion. We found the Hopscotch segregating at a frequency of 0.65; with a 0.4 frequency of homozygotes for the Hopscotch, a 0.5 frequency of heterozygotes, and a 0.1 frequency of homozygotes for the teosinte allele (no Hopscotch), showing no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and Hopscotch genotype we find a weak positive correlation between presence of the Hopscotch and tillering index on day 40 (p=0.0848), but no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and Hopscotch genotype, or culm diameter and Hopscotch genotype in phenotyping 1.

@@

\subsubsection*{Phenotyping 2}

For phenotyping 2 we measured tillering index every five days through day 50 for 302 plants. We followed the same transformations and analyses as described in phenotyping 1. We find the Hopscotch allele segregating at a frequency of 0.69, with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We find similar patterns as in phenotyping 1, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.

@@

In maize, alleles at important domestication loci (\emph{RAMOSA1}, \citep{SigmonVollbrecht2010}; \emph{barren stalk1}, \citep{Gallavotti2004}; and \emph{grassy tillers1}, \citep{Whipple2011}) have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \citep{Clark2006, Studer2011}. \citet{Studer2011} genotyped 90 accessions of teosinte (inbred and outbred), and provided the first evidence that the Hopscotch insertion is segregating in teosinte \citep{Studer2011}.

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by \citet{Studer2011} in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Crop to wild introgression has previously been observed at domestication loci (\citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}), but our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize) \citep{Hufford2013}.

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte. We do not have sympatric maize samples from many of our populations, however, so we cannot rule out the possibility that Hopscotch alleles in these teosinte populations are segregating in sympatric maize. Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1. The similarity in LD in the \emph{tb1} region in populations with and without the Hopscotch provides further support for an explanation other than introgression. Additionally, assignment to maize in this region in our STRUCTURE analysis is in fact lower than the average across chromosome 1 (\ref{Fig3Structure}, \ref{Table4Q}).

@@

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the Hopscotch in these populations. In the ancient introgression scenario, recombination would have broken up linked variants surrounding the Hopscotch and we would find LD patterns similar to the ones we report here. However, it seems unlikely that there would have been ancient introgression between these populations of teosinte and their sympatric maize populations and not present-day introgression. Studies have demonstrated the presence of ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \citep{Hufford2013, Ellstrand2007, vanHeerwaarden2011}.

Deviations from expected genotype frequencies under Hardy-Weinberg equilibrium at the Hopscotch insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus. Previous studies using both SSRs and genome-wide SNP data have provided evidence for a population bottleneck in the San Lorenzo population that has resulted in lower levels of diversity across the genome \citep{Hufford2010, Pyh\"aj\"arvi2013}. The lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D in the San Lorenzo population that we present here are consistent with these previous findings. Although, if the Hopscotch is segregating as neutral variation in teosinte, the probability that it would have simultaneously drifted to high frequency in multiple populations is quite low; however, we find no evidence of selection or introgression at or near the \emph{tb1} locus in our sample set. Interestingly, many of our populations with high frequency of the Hopscotch allele in the Jalisco region overlap with the Jalisco cluster in \citet{Fukunaga2005}.

\mbh{many populations in the Jalisco cluster of Fukunaga seem to have high frequency of the hopscotch allele.  Why?  Only ancient introgression in Jalisco and not in the Balsas? If hopscotch came from standing variation, is this more consistent with a Jalisco domestication?  I know, heresy!!  LV, can you look at the \emph{parviglumis} populations with high hopscotch and see if they are all in the Jalisco cluster of Fukunaga?  Might be a nice observation we could make here...could even go so far as saying the results from Matsuoka deserve another look.} \lev{Matt, yes they are in the Jalisco cluster, I added in a sentence mentioning it, but am not sure if that's the direction you were wanting me to take with it or not}

>>>>

\subsection*{Genotyping Analysis} \jri{please drop scripts for bayenv and STRUCTURE into github}

@@

Genotype of the Hopscotch insertion was confirmed with two PCRs for 837 individuals. Among the 247 maize landrace accessions genotyped, all but 8 were homozygous for the presence of the insertion (\ref{TableS1Mays}, \ref{TableS2}). \jri{please fix table/figure references to say table/figure} Within our \emph{parviglumis} and \emph{mexicana} samples we found the Hopscotch insertion segregating in 37 and 4 populations \jri{this is confusing as the map shows <37 populations. we should be consistent about reporting for all pops, or all pops with n>8, etc. or at least be explicit what cutoff we are using for each result}, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico in both subspecies (\ref{Fig1Map}) \jri{the map only shows parviglumis. can we add mexicana? the text makes it seem as if both should be on the map}. We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p<0.05).  \jri{in what direction? too many hets? what's the F? }

\jri{please fix whitespace and black border on figure}
%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=250mm]{Fig1Map.pdf}
    \caption{Map showing the frequency of the Hopscotch allele in populations of \emph{Zea mays} ssp. \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of alleles sampled. } 
\label{Fig1Map}
  \end{center}
\end{figure*}
%--------------------------------------------

Using our Hopscotch genotyping data, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled 8 or more alleles. F\textsubscript{CT} is 0 within our dataset, and we found similar levels of F\textsubscript{ST} among populations within each subspecies (0.22) \jri{is this an average? this comes from libsequence?} and among all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi2013} (\ref{Table1Fst}). \jri{the table caption says just sequence data, but it has hopscotch genotypes too}
% Table 1
\begin{table}[htbp]
  \centering
  \caption{Pairwise F\textsubscript{CT} values from sequence data}
    \begin{tabular}{lccc}
 %   \toprule
    \multicolumn{1}{c}{\textbf{Comparison}} & \textbf{Seq. Region 2} & \textbf{Seq.  Region 1} & \textbf{Hopscotch} \\
 %   \midrule
    EjuA \& EJuB & 0     & 0     & 0 \\
    EjuA \& MSA & 0.328 & 0.326 & 0.186 \\
    EjuA \& SLO & 0.258 & 0.416 & 0.28 \\
    EjuB \& MSA & 0.365 & 0.397 & 0.188 \\
    EjuB \& SLO & 0.29  & 0.512 & 0.28 \\
    MSA \& SLO & 0     & 0.007 & 0.016 \\
%    \bottomrule
    \end{tabular}
  \label{Table1Fst}
\end{table}
Although we found large variation in Hopscotch allele frequency among our populations, BayEnv analysis did not indicate a correlation between the Hopscotch insertion and environmental variables (all Bayes Factors $<$ 1; \ref{TableS3BayEnv}).

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A has the highest values of haplotype diversity, $\theta_W$, and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is more positive in San Lorenzo (\ref{Table2Diversity}). \jri{can drop theta W from table and text. we show pi and D, which is sufficient. fix caption in table. also add Hopscotch allele frequencies to table or list somewhere -- i think it's useful for comparison of seq stats }
% Table 2 Diversity Statistics
\begin{table}[htbp]
  \centering
  \caption{Add caption}
    \begin{tabular}{lccccc}
%    \toprule
    Population & \# Haplotypes & Hap. Diversity & $\hat\theta_W$    & $\hat\theta_\pi$    & Tajima's D \\
%    \midrule
    \multicolumn{6}{c}{\textit{Seq. region 2 (50kb upstream)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.89394 & 0.01548 & 0.01763 & 0.6231 \\
    \multicolumn{1}{c}{EJUB} & 8     & 0.89394 & 0.01493 & 0.01591 & 0.29504 \\
    \multicolumn{1}{c}{MSA} & 3     & 0.68182 & 0.01111 & 0.01055 & -0.22212 \\
    \multicolumn{1}{c}{SLO} & 4     & 0.74242 & 0.01167 & 0.01413 & 0.93185 \\
    \multicolumn{6}{c}{\textit{Seq.  region 1(5? UTR)}} \\
    \multicolumn{1}{c}{EJUA} & 8     & 0.85897 & 0.00874 & 0.00527 & -1.64955 \\
    \multicolumn{1}{c}{EJUB} & 5     & 0.70909 & 0.00663 & 0.00378 & -1.83123 \\
    \multicolumn{1}{c}{MSA} & 6     & 0.68182 & 0.00646 & 0.00373 & -1.75506 \\
    \multicolumn{1}{c}{SLO} & 3     & 0.31818 & 0.00176 & 0.00137 & -0.72873 \\
 %   \bottomrule
    \end{tabular}
  \label{Table2Diversity}
\end{table}

For region 2, haplotype diversity, $\theta_W$, and $\theta_\pi$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (\ref{Table2Diversity}). Tajima's D is positive in all populations except San Lorenzo, indicating an excess of low frequency variants in this population (\ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are 0 for both sequenced regions as well as for the Hopscotch, while they are high for other population pairs (\ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILs; data from Maize HapMapV2, \citep{Chia2012}) do not reveal any clear clustering pattern with respect to population or Hopscotch genotype (\ref{FigureS3NJ}); individuals within our sample that have the Hopscotch insertion do not group with the teosinte inbred lines or the lines of domesticated maize that have the Hopscotch insertion.

@@

The teosinte populations \jri{which?} with the highest frequency of the Hopscotch insertion in this study were sympatric with cultivated maize. Our initial hypothesis was that the high frequency of the Hopscotch element in these populations could be attributed to introgression from maize into teosinte. To investigate this possibility we examined overall patterns of linkage disequilibrium across chromosome one, and specifically in the \emph{tb1} region. If the Hopscotch is found in these populations due to recent introgression we would expect to find large blocks of linked markers near this element. We find no evidence of elevated linkage disequilibrium between the Hopscotch and SNPs surrounding the \emph{tb1} region in our resequenced populations (\ref{Fig2LD}), and $r^{2}$ in the \emph{tb1} region  does not differ significantly between populations with (average $r^{2}$ of 0.085) and without the Hopscotch genotype (average $r^{2}=0.082$). In fact, average $r^{2}$ is lower in the \emph{tb1} region ($r^{2}=0.056$) than across the rest of chromosome 1 ($r^{2}=0.083$) (\ref{Table3R2}). \jri{table is too wide, need to round numbers, and column headers are messed up.}

% Table 3 $r^{2}$ values
\begin{table}[htbp]
  \centering
  \caption{$r^{2}$ values between SNPs in the \emph{tb1} region (positions 264,596,664-265,891,456 on chromosome 1 of the maize AGPv2 genome) and the rest of chromosome 1, within the 5' UTR (Sequenced region 1), and within the 66,169 bp upstream region (Sequenced region 2).}
    \begin{tabular}{lccrc}
 %   \toprule
    \multirow{2}[2]{*}{Population} & \multirow{2}[2]{*}{Chromosome 1} & \multirow{2}[2]{*}{tb1 region} & \multicolumn{1}{c}{\multirow{2}[2]{*}{Seq. region 1}} & \multirow{2}[2]{*}{Seq. region 1 } \\
 %   \midrule
          &       &       & \multicolumn{1}{c}{} &  \\
    Ejutla A & 0.095426101 & 0.050304 & 0.747295 & 0.214933 \\
    Ejutla B & 0.068681837 & 0.051295 & 0.660354 & 0.186395 \\
    La Mesa & 0.069500533 & 0.053306 & 0.914286 & 0.766234 \\
    San Lorenzo & 0.100536784 & 0.067251 & 0.912281 & 0.636364 \\
    \bottomrule
    \end{tabular}
  \label{Table3R2}
\end{table}

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig2LDPlots.pdf}
    \caption{Linkage disequilibrium for SNPs in Mb 261-268 on chromosome 1. The yellow rectangle indicates the location of the Hopscotch insertion and the green represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D). San Lorenzo} 
\label{Fig2LD}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

The lack of clustering of Hopscotch genotypes in our NJ tree as well as the lack of LD around \emph{tb1} does not support the hypothesis that the Hopscotch insertion in these populations of \emph{parviglumis} is the result of recent introgression. However, to further explore this hypothesis we performed a STRUCTURE analysis using Illumina MaizeSNP50 data from four of our \emph{parviglumis} populations (EjuA, EjuB, MSA, and SLO) and the maize 282 diversity panel \citep{Cook2012, Phyh\"aj\"arvi2013}. The linkage model implemented in STRUCTURE can be used to identify ancestry of blocks of linked variants, which would arise as a result of recent admixture between populations. If the Hopscotch insertion is present in populations of \emph{parviglumis} as a result of recent admixture with domesticated maize, we would expect the insertion and linked variants in surrounding sites to be assigned to the "maize" cluster in our STRUCTURE runs, not the "teosinte" cluster. In all runs, assignment to maize in the \emph{tb1} region across all four \emph{parviglumis} populations is low (average 0.017) \jri{is this really 0.017 or 0.17?} and much below the chromosome-wide average (0.20; \ref{Fig3Structure}).

% Figure 3 STRUCTURE Plot 4 Pops
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig3Structure.pdf}
    \caption{STRUCTURE assignment to maize across a section of chromosome 1. The dotted lines mark the beginning of the sequenced region 50kb upstream (Sequenced region 2) and the end of the \emph{tb1} ORF. } 
\label{Fig3Structure}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------
\jri{please put figures in the text rather than at the end.}

@@

\subsubsection*{Phenotyping}

To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency \jri{(X)} of the Hopscotch insertion from our initial genotyping. We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants from 23 sampling locations within the San Lorenzo population, and genotyped plants for the Hopscotch insertion. \jri{the 23 sampling locales can go in methods and scratch from here}. We found the Hopscotch segregating at a frequency of 0.65 with no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and Hopscotch genotype we find a weak positive correlation between presence of the Hopscotch and tillering index on day 40 (p=0.0848), but no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and Hopscotch genotype, or culm diameter and Hopscotch genotype in phenotyping 1. \jri{shouldn't we expect a negative correlation betweeh Hop and TI on day 40? need to have an A and B in the figure and explain one is for pheno1 and one is for pheno2. please explain whiskers and dots on figure too.}

@@

We performed a second grow-out  of teosinte to assess whether lighting conditions or sample size may have affected our ability to detect and effect of \emph{tb1}.  For the second grow-out we measured tillering index every five days through day 50 for 302 plants. We found the Hopscotch allele segregating at a frequency of 0.69, \jri{does this violate HWE? didn't we choose numbers of each genotype here, so that HWE isn't appropriate? if so, probably should drop genotype frequency data here} with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We found similar patterns, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.

@@

In maize, alleles at important domestication loci (\emph{RAMOSA1}, \citep{SigmonVollbrecht2010}; \emph{barren stalk1}, \citep{Gallavotti2004}; and \emph{grassy tillers1}, \citep{Whipple2011}) have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \citep{Clark2006, Studer2011}. \citet{Studer2011} genotyped 90 accessions of teosinte (inbred and outbred), providing the first evidence that the Hopscotch insertion is segregating in teosinte \citep{Studer2011}.

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by \citet{Studer2011} in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the Hopscotch allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus than in the Balsas region where maize was domesticated \jri{cite matsuoka} . While gene flow from crops into their wild relatives is well-known, (\citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}),\jri{do all these show evidence of introgression AT domestication loci, or just introgression in general?}  our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize) \citep{Hufford2013}.

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte.  Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1, and assignment to maize in this region in our STRUCTURE analysis is lower than the average across chromosome 1 (\ref{Fig3Structure}, \ref{Table4Q}). Together, these data point to an explanation other than recent introgression for the high observed frequency of Hopscotch in some of  our \emph{parviglumis} populations.

@@

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the Hopscotch in these populations. If the Hopscotch allele was introgressed in the distant past, they could have been sufficient recombination to break up any initial LD, leading to observations similar to those obtained here.  We find this scenario less plausible, however, as there is no reason why gene flow should have been high in the past but absent in present-day sympatric populations.  In fact, early generation maize-teosinte hybrids are easy to find in these populations today (MB Hufford, pers. observation), and genetic data support ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \citep{Hufford2013, Ellstrand2007, vanHeerwaarden2011}.

Other explanation for differential frequencies of the Hopscotch among teosinte populations include both drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyh\"aj\"arvi2013}, and the lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D we present here are consistent with these findings. \jri{deviations from HWE may be consistent too if we see excess of homozygotes. do we?}. This bottleneck, however, does not explain differences in Hopscotch allele frequency among populations, and the available information on diversity and population structure among these populations \citep{Hufford2010, Pyh\"aj\"arvi2013} is not suggestive of colonization or other demographic events that might predict a high frequency of the allele in multiple populations.  \jri{here we need a few sentences on selection.  the 5' UTR has much more negative D than the upstream. do we know the Hop genotype for sequenced lines? can we separate the sequences into hop/no hop and look for differences? it wasn't until we did this that gt1 stuff really popped out.}
%   from expected genotype frequencies under Hardy-Weinberg equilibrium at the Hopscotch insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus..

%\mbh{many populations in the Jalisco cluster of Fukunaga seem to have high frequency of the hopscotch allele.  Why?  Only ancient introgression in Jalisco and not in the Balsas? If hopscotch came from standing variation, is this more consistent with a Jalisco domestication?  I know, heresy!!  LV, can you look at the \emph{parviglumis} populations with high hopscotch and see if they are all in the Jalisco cluster of Fukunaga?  Might be a nice observation we could make here...could even go so far as saying the results from Matsuoka deserve another look.} \lev{Matt, yes they are in the Jalisco cluster, I added in a sentence mentioning it, but am not sure if that's the direction you were wanting me to take with it or not}
